---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mixtime

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![R build status](https://github.com/mitchelloharawild/mixtime/workflows/R-CMD-check/badge.svg)](https://github.com/mitchelloharawild/mixtime/actions)
[![Coverage status](https://codecov.io/gh/mitchelloharawild/mixtime/branch/master/graph/badge.svg)](https://codecov.io/github/mitchelloharawild/mixtime?branch=master)
<!-- badges: end -->

mixtime provides flexible time classes for representing mixed temporal granularities
and custom calendar structures.

The feature wishlist for this package includes:

* Multiple temporal granularities (sub-daily, daily, weekly, monthly, quarterly, etc.)
* Optional and custom temporal origins
* Custom calendar structures (working days/hours, holiday effects, trading days)
* Mixed temporal classes (daily and weekly granularities in same vector)

Stretch goals for the package include:

* Representing temporal nesting (2020-01-05 is nested by 2020-01 is nested by 2020)
* `time_join()` operation which respects temporal nesting

## Installation

<!-- You can install the released version of mixtime from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("mixtime") -->
<!-- ``` -->

The development version can be installed from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mitchelloharawild/mixtime")
```
## Example

Creating mixtimes.

```{r demo}
library(mixtime)
yearmonth(0:11) # By default, time classes have an origin
yearmonth(0:11) - yearmonth(0) # However some operations can produce mixtimes without origins

yearquarter(0:3)
yearquarter(0:3) - yearquarter(0)

# Different temporal granularities can be combined:
c(yearquarter(0:3), yearmonth(0:11))
```
Sequences of mixtimes.

```{r sequence}
seq(yearmonth(0), yearmonth(10), by = 1)
```

Integration with tsibble.

```{r tsibble}
tsibble::tsibble(time = yearmonth(0:5), index = time)
tsibble::tsibble(time = c(year(0), yearquarter(0:3), yearmonth(0:5), Sys.Date()), index = time)
```

Change granularities by updating the calendar.

```{r granularity}
x <- yearmonth(0:11)
tsibble::tsibble(
  yearmonth = x,
  yearquarter = set_time_units(x, tu_quarter(1)),
  year = set_time_units(x, tu_year(1)),
  index = yearmonth
)
```
