---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mixtime <a href="https://pkg.mitchelloharawild.com/mixtime/"><img src="man/figures/logo.svg" align="right" height="139" alt="mixtime website" /></a>

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![R-CMD-check](https://github.com/mitchelloharawild/mixtime/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mitchelloharawild/mixtime/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

mixtime provides flexible time classes for time series analysis and forecasting with mixed temporal granularities. It is designed for handling temporal data at different frequencies, making it ideal for:

* Changing observation frequencies (e.g., historical quarterly records now reported monthly)
* Analysis involving data at different temporal resolutions (e.g., quarterly tourism and monthly turnover)
* Forecasting with temporal reconciliation across multiple time scales

## Key Features

**📈 Linear Time** - Create linear time vectors with `linear_time()` or with helpers:

  * `yearquarter()`, `yearmonth()`, `yearmonthday()`, `yearweek()`

**🔄 Cyclical Time** - Create cyclical time vectors with `cyclical_time()` or with helpers:

  * `month_of_year()`, `day_of_year()`, `day_of_month()`, `time_of_day()`, `day_of_week()`, `week_of_year()`

**🕰️ Time types**

  * Discrete and continuous time models
  * Timezone support for all chronons
  * Custom granule sizes (e.g. `week(2L)` for fortnights)
  * Mixed granularity vectors for different temporal resolution over time (or series)

**📅 Calendar Systems** - Support for several calendars:

  * `cal_gregorian` - Gregorian dates (e.g. 2026-02-17)
  * `cal_isoweek` - ISO week dates (e.g. 2026-W8-Tue)
  * More calendars coming, including custom censored calendars

**🧮 Temporal Operations**

  * Rounding: `floor_time()`, `round_time()`, `ceiling_time()`
  * Sequencing: `seq()` for linear and cyclical time points

## Installation

The development version can be installed from [GitHub](https://github.com/mitchelloharawild/mixtime) with:

``` r
# install.packages("remotes")
remotes::install_github("mitchelloharawild/mixtime")
```

## Usage

```{r}
library(mixtime)
demo_time <- as.POSIXct("2026-02-22 18:30:42", tz = "UTC")
demo_date <- as.Date("2026-02-22")
```

The mixtime package is designed around the time units of calendars. These time units are used to create and manipulate time vectors.

### Calendars and time units

Calendars have a `cal_*` prefix, which contain a set of time units that can be accessed with `<cal>$<unit>`.

```{r}
# The Gregorian calendar
cal_gregorian

# A 1-month time unit
cal_gregorian$month(1L) # (1L is integer 1)

# A 2-week time unit (fortnights)
cal_isoweek$week(2L)
```

### Linear Time

A linear time vector uses time units to define the resolution of time points, known as a `chronon` (the smallest unit of time). When the input time has a different resolution than the `chronon`, it will be automatically converted.

```{r}
linear_time(demo_date, chronon = cal_gregorian$month(1L))
```

Discrete time models (integer-based values) are used by default, however continuous time models (double-based values) can be used with `discrete = FALSE` to allow fractional chronons.

```{r}
# February 22nd is 75% through the month (in non-leap years)
linear_time(demo_date, chronon = cal_gregorian$month(1L), discrete = FALSE)
```

Linear time helper functions are available to quickly create common time points.

```{r linear-time}
# Create time vectors at different granularities
yearquarter(demo_date) + 0:7
yearmonth(demo_date) + 0:11
yearweek(demo_date) + 0:10
yearmonthday(demo_date) + 0:6
```

The mixtime package allows time of different granulities to be combined in a single vector.

```{r}
c(
  year(demo_date), yearquarter(demo_date), 
  yearmonth(demo_date), yearweek(demo_date)
)
```

### Cyclical Time

A cyclical time vector is defined by two calendar time units: a `chronon` (the smaller unit of time) and a `cycle` (the larger unit that the `chronon` loops over).

```{r}
# The `calendar` argument provides a masking scope to `chronon` and `cycle`
cyclical_time(demo_date, chronon = day(1L), cycles = list(week(1L)), calendar = cal_isoweek)
```

There are several cyclical time helper functions for convenience.

```{r cyclical-time}
# Extract cyclical components
month_of_year(demo_date)
week_of_year(demo_date)
day_of_week(demo_date)

# Continuous cyclical time shows progress through chronons
day_of_week(demo_time, discrete = FALSE)
```

### Timezones

All linear and cyclical time vectors support timezones via the `tz` argument.

```{r timezone}
demo_time
# Same day (Sunday) in LA
yearmonthday(demo_time, tz = "America/Los_Angeles")
yearmonthday(demo_time, tz = "America/Los_Angeles", discrete = FALSE)
day_of_week(demo_time, tz = "America/Los_Angeles")

# Next day (Monday) in Melbourne (23% through the 23rd)
yearmonthday(demo_time, tz = "Australia/Melbourne")
yearmonthday(demo_time, tz = "Australia/Melbourne", discrete = FALSE)
day_of_week(demo_time, tz = "Australia/Melbourne")
```

### Temporal Manipulation

Linear time points can be adjusted to the floor, ceiling, or rounded to a specified time unit.

```{r manipulation}
# Round dates to different granularities
floor_time(demo_date, cal_gregorian$month(1L))
round_time(demo_date, cal_isoweek$week(1L))
ceiling_time(demo_date, cal_gregorian$month(1L))
```

### Time Sequences

The `seq()` function creates sequences of time points iterating by a given time unit.

```{r sequences}
# Integer increments (advances by chronon's natural unit)
seq(yearmonth(demo_date), by = 1L, length.out = 10)

# Calendar time units allow sequencing by other units
seq(yearmonthday(demo_date), by = cal_gregorian$month(1L), length.out = 8)
```


