---
title: "Extending mixtime"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{extending-mixtime}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mixtime)
```

This vignette documents the extension points made available in mixtime for creating new calendars and time units. As a running example, we will build a financial calendar where a year has 13 periods of exactly 4 weeks each (a "4-4-5" style retail/fiscal calendar simplified to equal periods).

The extension points covered are:

* Calendars: `new_calendar()`
* Time units: S7 classes inheriting from `mt_unit`, `mt_tz_unit`, or `mt_loc_unit`
* Calendar arithmetic: `chronon_cardinality()`, `chronon_divmod()`, and `chronon_epoch()`
* Labels: 
  * Time units: `time_unit_abbr()`, `time_unit_full()`
  * Time formats: `chronon_format_linear()`, `chronon_format_cyclical()` 
  * Granule labels: `linear_labels()`, `cyclical_labels()`
* Time vector helpers: `new_linear_time_fn()` and `new_cyclical_time_fn()`

The mixtime package uses S7 for all extension methods. If you are new to S7, the vignette on [generics and methods](https://rconsortium.github.io/S7/articles/generics-methods.html) is a good place to start. It is best to develop new calendars and time units in a separate package that imports mixtime, and registering your methods with the following [`.onLoad` hook](https://rconsortium.github.io/S7/articles/packages.html):

```{r}
.onLoad <- function(...) {
  S7::methods_register()
}
```

## Time units

Time units are the building blocks of calendars. Time units are S7 classes that inherit `mt_unit`, a purely abstract time unit class.

In practice, most time units should inherit either:

* `mt_tz_unit` for civil-time units whose boundaries depend on a timezone (e.g. months, days, hours, ...), or
* `mt_loc_unit` for astronomical-time units whose boundaries depend on geographic location (e.g. solar days).

All time units accept a numeric `.data` argument representing the size of the unit (e.g. `1L` for "one month"). Additional arguments needed for your time unit can be defined with S7 by:

* inheriting properties from `mt_tz_unit` (`tz` IANA names) or `mt_loc_unit` (for `lon`, `lat`, and `alt`),
* defining custom S7 properties with `S7::new_class(properties = ...)`.

The name of a time unit's S7 class becomes the internal identifier for that unit, so it should be unique across all calendars. The class name is not important to users of the calendar, but it should be descriptive for developers.

```{r time-units}
# A timezone-aware fiscal period and year units
S7::new_class("tu_fiscal_period", parent = mt_tz_unit)
S7::new_class("tu_fiscal_year", parent = mt_tz_unit)
```

These time units are typically exported within a calendar object (shown below) and accessed via `$` notation.

## Calendars

A calendar is a named collection of time units, created with `new_calendar()`. An optional `inherit` argument pulls in all units from an existing calendar (units defined in `...` take precedence). It is useful to inherit from a base calendar that defines common time units for civil and astronomical calendars. The base calendars provided by mixtime are:

* `cal_time_civil_midnight`: civil time units anchored at midnight (e.g. days, hours, minutes, seconds, ...)
* `cal_time_solar_***`: solar time units anchored at sunrise, noon, sunset or midnight (e.g. solar days)
* `cal_time_lunar`: lunar time units (e.g. lunar months, lunar phases)

```{r new-calendar}
cal_fiscal <- new_calendar(
  year   = S7::new_class("tu_fiscal_year", parent = mt_tz_unit),
  quarter = S7::new_class("tu_fiscal_quarter", parent = mt_tz_unit),
  month = S7::new_class("tu_fiscal_month", parent = mt_tz_unit),
  # Re-use ISO week units since they are equivalent
  week = cal_isoweek$week,
  # Inherit civil-time units (day, hour, minute, second, ...)
  inherit = cal_time_civil_midnight,
  class = "cal_fiscal"
)

cal_fiscal
```

Unit constructors are now accessible via `$`:

```{r calendar-access}
cal_fiscal$year(1L)
cal_fiscal$month(1L, tz = "UTC")
```

These time units can be used as chronons/cycles within `mixtime()` (and also `linear_time()`, `cyclical_time()`). The temporal relationships between these units must be defined by the calendar arithmetic methods described below.

## Calendar arithmetic

Two S7 generics drive all calendar calculations: `chronon_cardinality()` and `chronon_divmod()`. These functions are used to manipulate time, such as converting between units, generating sequences, and comparing time points. The details of how they work are not important for users of the calendar, but they are the main extension points for developers creating new calendars.

For a time unit to be useful in mixtime, at minimum a `chronon_cardinality()` method between itself and the next coarser unit must be defined. Time units with irregular relationships (e.g. days → months) also require `chronon_divmod()` to efficiently compute relationships between units.

Defining these methods for time units between different calendars is also possible, which allows for interoperability between calendars. In most cases, calendars will share a common base calendar (e.g. `cal_time_civil_midnight`) which provides a common time unit (e.g. civil days) or methods for converting between different calendars (e.g. civil midnight to solar midnight).

### Cardinality

The cardinality is the number of finer units that fit inside a coarser unit. For example, there are 7 days in a week, so the cardinality of `cal_isoweek$day(1L)` in terms of `cal_isoweek$week(1L)` is 7.

```{r}
chronon_cardinality(cal_isoweek$day(1L), cal_isoweek$week(1L))
```

For units with a fixed relationship (e.g. 7 days per week) the `at` argument can be ignored. The `at` variable disambiguates the point in time for variable cardinalities (e.g. days per month), `at` is the internal numeric representation of time in the coarser unit (e.g. months since epoch for days → months).

```{r}
chronon_cardinality(cal_gregorian$day(1L), cal_gregorian$month(1L), at = 0L) # Jan 1970
chronon_cardinality(cal_gregorian$day(1L), cal_gregorian$month(1L), at = 1L) # Feb 1970
chronon_cardinality(cal_gregorian$day(1L), cal_gregorian$month(1L), at = 25L) # Feb 1972 (leap year)
```

In our 4-4-5 fiscal calendar every year has 4 quarters, and every quarter has 3 months.

```{r cardinality-fixed}
# Each fiscal year is 4 quarters
S7::method(chronon_cardinality, list(cal_fiscal$quarter, cal_fiscal$year)) <-
  function(x, y, at = NULL) {
    vctrs::vec_data(y) * 4L / vctrs::vec_data(x)
  }
chronon_cardinality(cal_fiscal$quarter(1L), cal_fiscal$year(1L))

# Each fiscal quarter is 3 months
S7::method(chronon_cardinality, list(cal_fiscal$month, cal_fiscal$quarter)) <-
  function(x, y, at = NULL) {
    vctrs::vec_data(y) * 3L / vctrs::vec_data(x)
  }
chronon_cardinality(cal_fiscal$month(1L), cal_fiscal$quarter(1L))
```

The number of weeks in each month varies according to the 4-4-5 pattern. The `circsum()` helper function is useful for finding the cardinality of cyclical units like this (especially when the units are not length 1, e.g. the number of weeks in 2 fiscal months).

```{r}
S7::method(chronon_cardinality, list(cal_fiscal$week, cal_fiscal$month)) <-
  function(x, y, at = NULL) {
    # The number of weeks in each n-month period
    month_size <- vctrs::vec_data(y)
    nweeks <- circsum(c(4L, 4L, 5L), month_size)

    # Find which n-month period we're in based on the "at" position (months since epoch)
    period <- at %% length(nweeks) + 1L

    # Compute per-at cardinality before applying leap-week adjustment
    nweeks_result <- nweeks[period]

    # 53-week years add a 6th week in the last month of the year (4-4-6).
    # For a fiscal calendar with years starting on the first Monday of the year,
    # this happens when Jan 1 is Monday or also Sunday in a leap year.

    # Since 4-4-6 only affects the last month of the year, find periods containing
    # the last month of the year based on the "at" position (months since epoch)
    m1 <- at * month_size
    contains_dec <- which((m1 %% 12L) >= (12L - month_size))

    # Jan 1 1970 was Thursday. Each common year shifts dow +1, each leap year +2.
    # Total shift = (year - 1970) + leaps_since_1970
    year <- 1969L + m1[contains_dec] %/% 12L
    y <- year - 1L
    leaps <- y %/% 4L - y %/% 100L + y %/% 400L - 477L
    dow_jan1 <- ((year - 1970L) + leaps + 3L) %% 7L
    is_leap_year <- (year %% 4L == 0L & year %% 100L != 0L) | (year %% 400L == 0L)

    # 0-indexed day of week for Jan 1, where 0 = Monday, 1 = Tuesday, ..., 6 = Sunday
    has_53_weeks <- dow_jan1 == 0L | (dow_jan1 == 6L & is_leap_year)

    # Add the extra week to December of 53-week years
    nweeks_result[contains_dec[has_53_weeks]] <- nweeks_result[contains_dec[has_53_weeks]] + 1L

    # Scale by the number of weeks in the week time unit
    nweeks_result / vctrs::vec_data(x)
  }

# The 4-4-5 cycle for an entire fiscal year
chronon_cardinality(cal_fiscal$week(1L), cal_fiscal$month(1L), at = 0:11)

# 1973 is a 53-week year (starts on Monday), so December has 6 weeks
chronon_cardinality(cal_fiscal$week(1L), cal_fiscal$month(1L), at = 36:47)

# 2012 is a 53-week year (starts on Sunday in a leap year), so December has 6 weeks
chronon_cardinality(cal_fiscal$week(1L), cal_fiscal$month(1L), at = 504:515)

# The number of weeks in a multi-month period is the sum of the weeks in each month
chronon_cardinality(cal_fiscal$week(1L), cal_fiscal$month(2L), at = 0:5)
```

Cardinality methods are only required for adjacent time units (e.g. month → quarter, but not month → year), but they can be defined for non-adjacent units as an optimization. If a direct cardinality method is not defined, mixtime will attempt to derive by computing cardinality along a path of intermediate units.

```{r}
# The number of weeks in a fiscal year (derived from week → month → quarter → year)
chronon_cardinality(cal_fiscal$week(1L), cal_fiscal$year(1L), at = 0:4)

# The number of days in a fiscal year (derived from day → week → month → year)
chronon_cardinality(cal_fiscal$day(1L), cal_fiscal$year(1L), at = 0:4)
```

The cardinality methods also define which units are finer/coarser for operations requiring graph traversal, so it is essential that the first unit is the finer unit and the second unit is the coarser unit. The inverse cardinality method (with the units flipped) is automatically derived, so you should only define one direction (finer → coarser) of the relationship.

```{r}
# The number of fiscal years in a fiscal quarter
chronon_cardinality(cal_fiscal$year(1L), cal_fiscal$quarter(1L))
```

### Divmod

The divmod operation is a combined division and modulus that converts between time units and is defined with `chronon_divmod()` methods. These methods are required to efficiently convert between units with irregular relationships, such as days → months. For regular relationships (e.g. days → weeks) the divmod is automatically derived from `chronon_cardinality()` methods alone.

`chronon_divmod(from, to, x)` converts times `x` in the `from` units into time points in `to` units. For example, the date "1970-02-15" is day 45 since epoch (`x = 45` and `from = cal_gregorian$day(1L)`), converted to months (`to = cal_gregorian$month(1L)`) gives 1 month and 14 days (div = 1, mod = 14).

```{r}
chronon_divmod(cal_gregorian$day(1L), cal_gregorian$month(1L), 45L)
```

Methods for `chronon_divmod()` return a list with:

* `div` - the quotient (chronons in `to` units)
* `mod` - the remainder (leftover `from` chronons)

Note that all time values are zero-indexed, so `$div = 0` is January 1970, and `$mod = 0` is the first day of the month. So the above result can be interpreted as `$div = 1` is February 1970 and `$mod = 14` is 14 remaining days (i.e. the 15th).

The fiscal calendar has an irregular cardinality for weeks → months, so we need to define divmod methods for days → months. Additional methods are optional (e.g. days → years) but can be defined for efficiency.

```{r divmod}
# Divmod for converting weeks → months
S7::method(chronon_divmod, list(cal_fiscal$week, cal_fiscal$month)) <-
  function(from, to, x) {
    # This implementation assumes n-month periods nest perfectly within years for simplicity
    # c(1, 2, 3, 4, 6, 12) month sizes are acceptable
    if (!vctrs::vec_data(to) %in% c(1L, 2L, 3L, 4L, 6L, 12L)) {
      stop("Only month sizes of 1, 2, 3, 4, 6, or 12 are supported", call. = FALSE)
    }

    # The number of weeks in each n-month period
    weeks_len <- circsum(c(4L, 4L, 5L), vctrs::vec_data(to))

    # The total weeks in a full n-month cycle
    weeks_tot <- sum(weeks_len)

    # Number of 53rd weeks included in `x`
    279 # 53rd weeks from year 0 to 1970.
    
    # 53rd weeks in 400-year cycle
    71 * x %/% 400L +

    # 53rd weeks in remaining years, which depends on the position within the cycle

    # Find which n-month cycle we're in based on `x` (weeks since epoch)
    period_full <- (x %/% weeks_tot)

    # Find which part within the n-month cycle we're in
    weeks_seq <- cumsum(weeks_len[-length(weeks_len)])
    period_part <- rowSums(outer(x %% weeks_tot, weeks_seq, ">="))

    list(
      # The number of whole n-months (total cycles + part cycles) is the div result
      div = period_full * length(weeks_len) + period_part,
  
      # The number of leftover weeks is the number of weeks into the current n-month period
      mod = x %% weeks_tot - c(0L, weeks_seq)[period_part + 1L]
    )
  }

# Week 20 of the fiscal calendar is month 4 (May) with 3 additional weeks remaining
with(cal_fiscal, chronon_divmod(week(1L), month(1L), 19L))
```

If a direct `chronon_divmod()` method is not defined for a pair of units, mixtime will attempt to derive it by chaining together divmod operations along the shortest path of intermediate units. 

```{r}
# Divmod for converting days → years is derived from days → weeks → months → years
# Day 839 since epoch (1972-04-19) is fiscal year 2 with 107 additional days
with(cal_fiscal, chronon_divmod(day(1L), year(1L), 839L))
```

### Epoch anchors

The `chronon_epoch()` method defines the anchor point for a time unit, which is the reference point for all time calculations. The epoch is used as the zero point for time points, for example time point 0 in `cal_gregorian$year(1L)` corresponds to the year 1970. This anchor maps the internal numeric representation of time (e.g. years since epoch) to a real-world time point (e.g. "1970"), and is used as the origin for numeric inputs and labelling linear parts of time.

The years in our fiscal calendar are anchored to the same epoch as the Gregorian calendar, so `cal_fiscal$year(1L)` also needs an epoch of 1970.

```{r}
S7::method(chronon_epoch, cal_fiscal$year) <- function(x) 1970L
```

## Labels

The label methods control how time units and time points are displayed as text. They are used in messages, plot axes, and when formatting time vectors with `format()`. The formatting methods return format strings that are parsed by `format_mixtime()` to generate the final labels.

### Time units

Each time unit has two format methods:

* `time_unit_full()` returns the singular full name of the unit (e.g. "fiscal year")
* `time_unit_abbr()` returns the short abbreviation of the unit (e.g. "FY")

The full name is used in messages (e.g. "Can't convert from fiscal year to day") and the abbreviation is used for displaying time intervals and durations.

```{r}
S7::method(time_unit_full, cal_fiscal$year) <- function(x) "fiscal year"
S7::method(time_unit_abbr, cal_fiscal$year) <- function(x) "FY"
S7::method(time_unit_full, cal_fiscal$quarter) <- function(x) "fiscal quarter"
S7::method(time_unit_abbr, cal_fiscal$quarter) <- function(x) "FQ"
S7::method(time_unit_full, cal_fiscal$month) <- function(x) "fiscal month"
S7::method(time_unit_abbr, cal_fiscal$month) <- function(x) "FM"
```

The time unit abbreviations are also used in the default time formatting string, so we can now create and view our fiscal time vector:

```{r}
linear_time(as.Date("2025-02-15"), chronon = cal_fiscal$year(1L))
```

The linear time helper functions can also be used with the new calendar:

```{r}
year(as.Date("2025-02-15"), calendar = cal_fiscal)
```

Cyclical time vectors can also be created with the new calendar, and the default formatting string will use the time unit abbreviations:

```{r}
cyclical_time(as.Date("2025-02-15"), chronon = cal_fiscal$week(1L), cycle = cal_fiscal$year(1L))
month_of_year(as.Date("2025-02-15"), calendar = cal_fiscal)
```

### Time formats

<!-- TODO: Time formatting vignette -->
The default formatting for time vectors is defined by `chronon_format_linear()` for linear time and `chronon_format_cyclical()` for cyclical time. These methods return mixtime format strings which are used to format and print time vectors. A mixtime format string is a [glue](https://glue.tidyverse.org/) style string where time granules wrapped in `{}` are replaced with granule labels for the corresponding time points.

There are two types of granule labels useful for these time format strings:

* `lin(<time unit>)` - Granule labels for linear time units, usually the largest granule (e.g. `lin(year(1L))`).
* `cyc(<time unit>, <cycle unit>)` - Granule labels for cyclical time units, where the labels are relative to the cycle (e.g. `cyc(month(1L), year(1L))` is the month within the year).

The `<time unit>` and `<cycle unit>` placeholders in the format string must evaluate to a time unit of a calendar. For convenience, unevaluated named time units (e.g. `lin(year)`) default to a size 1 time unit (e.g. `lin(year(1L))`), and the calendar is inferred from the time vector being formatted. It is also possible to explicitly specify the calendar (e.g. `lin(cal_fiscal$year(1L))`), which allows time formats to use granules from different calendars.

Linear time formats dispatch on the chronon and calendar:
```{r}
S7::method(chronon_format_linear, list(cal_fiscal$year, S7::new_S3_class("cal_fiscal"))) <- function(x, cal) {
  paste0(time_unit_abbr(x), "{lin(year)}")
}
```

Cyclical time formats dispatch on the chronon and cycle:
```{r}
S7::method(chronon_format_cyclical, list(cal_fiscal$month, cal_fiscal$year)) <- function(x, y) {
  paste0(time_unit_abbr(x), "{cyc(month, year)}")
}
```


Linear granule labels are evaluated with `linear_labels()`, and cyclical granule labels are evaluated with `cyclical_labels()`. The default methods for these generics return the internal numeric representation of the time point (e.g. "0", "1", "2", ...), but they can be overridden to return custom labels (e.g. "FY1", "FY2", "FY3", ... for fiscal years).

### Granule labels

### `time_unit_full()` and `time_unit_abbr()`

These return the singular full name and the short abbreviation for a unit,
respectively. They are used in messages and in tsibble interval displays.


### `cyclical_labels()`

`cyclical_labels(granule, cycle, i)` returns human-readable labels for the
0-indexed position `i` of `granule` within `cycle`. This is used when
formatting cyclical time vectors (e.g. "Jan", "Feb", … for months within a
year).

The default method falls back to `paste0(time_unit_abbr(granule), i)`, so you
only need to override it when you want custom labels.
