---
title: "Extending mixtime"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{extending-mixtime}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mixtime)
```

This vignette documents the extension points made available in mixtime for
creating new calendars and time units. As a running example, we will build a
**financial calendar** where a year has 13 periods of exactly 4 weeks each
(a "4-4-5" style retail/fiscal calendar simplified to equal periods).

The topics covered are:

* Calendars: `new_calendar()`
* Time units: S7 classes inheriting from `mt_unit`, `mt_tz_unit`, or `mt_loc_unit`
* Calendar arithmetic: `chronon_cardinality()` and `chronon_divmod()`
* Labels: `time_unit_abbr()`, `time_unit_full()`, and `cyclical_labels()`
* Time vector helpers: `new_linear_time_fn()` and `new_cyclical_time_fn()`

## Time units

Time units are the building blocks of calendars. Every time unit is an S7
class that inherits from one of three base classes:

| Base class    | Properties | Use case |
|---------------|------------|----------|
| `mt_unit`     | none beyond `.data` | Purely abstract / epoch-relative units |
| `mt_tz_unit`  | `tz` (IANA timezone string) | Civil-time units aligned to a timezone |
| `mt_loc_unit` | `lat`, `lon`, `alt` | Astronomical units whose boundaries depend on geographic location (e.g. solar days) |

All three accept a single numeric `.data` argument representing the *size* of
the unit (e.g. `1L` for "one month"). Units are created with `S7::new_class()`:

```{r time-units}
# A timezone-aware fiscal period unit
tu_fiscal_period <- S7::new_class("tu_fiscal_period", parent = mt_tz_unit)

# A timezone-aware fiscal year unit
tu_fiscal_year <- S7::new_class("tu_fiscal_year", parent = mt_tz_unit)
```

Instantiate a unit by calling it like a function:

```{r unit-constructors}
tu_fiscal_period(1L)
tu_fiscal_period(1L, tz = "America/New_York")
```

## Calendars

A **calendar** is a named collection of time units. Use `new_calendar()` to
define one. Each argument name becomes a constructor accessible via `$`
notation on the returned calendar object. An optional `inherit` argument pulls
in all units from an existing calendar (units defined in `...` take
precedence).

```{r new-calendar}
cal_fiscal <- new_calendar(
  year   = tu_fiscal_year,
  period = tu_fiscal_period,
  # Inherit civil-time sub-day units (hour, minute, second, …) from the
  # built-in Gregorian calendar
  inherit = cal_gregorian,
  class = "cal_fiscal"
)

cal_fiscal
```

Unit constructors are now accessible via `$`:

```{r calendar-access}
cal_fiscal$year(1L)
cal_fiscal$period(1L, tz = "UTC")
```

## Calendar arithmetic

Two S7 generics drive all calendar calculations: `chronon_cardinality()` and
`chronon_divmod()`. You must implement at least one of them for every pair of
adjacent units in your calendar.

### Cardinality

`chronon_cardinality(from, to, at = NULL)` returns the number of `to` units
that fit inside one `from` unit. For units with a *fixed* relationship (e.g.
7 days per week) the `at` argument can be ignored. For *variable* relationships
(e.g. days per month) `at` is the 0-indexed count of `from` units since the
Unix epoch (1970-01-01 UTC) and should be used to compute the exact count.

In our fiscal calendar every period is exactly 4 weeks (28 days), and every
year contains exactly 13 periods - so both relationships are fixed:

```{r cardinality}
# 13 fiscal periods per fiscal year
S7::method(chronon_cardinality, list(cal_fiscal$year, cal_fiscal$period)) <-
  function(x, y, at = NULL) {
    vec_data(x) * 13L / vec_data(y)
  }

# Each fiscal period is 28 days (4 weeks × 7 days)
S7::method(chronon_cardinality, list(cal_fiscal$period, cal_gregorian$day)) <-
  function(x, y, at = NULL) {
    vec_data(x) * 28L / vec_data(y)
  }

# Verify
chronon_cardinality(cal_fiscal$year(1L),   cal_fiscal$period(1L)) # 13
chronon_cardinality(cal_fiscal$period(1L), cal_gregorian$day(1L))  # 28
```

### Divmod

`chronon_divmod(from, to, x)` converts an integer vector `x` (chronons
measured in `from` units since the epoch) into `to` units. It returns a list
with:

* `div` - the quotient (chronons in `to` units)
* `mod` - the remainder (leftover `from` chronons)

This is needed for irregular conversions that cannot be derived from
`chronon_cardinality()` alone, such as mapping a day number to its containing
month. For our fiscal calendar the epoch anchor is 1970-01-01, which falls in
fiscal period 1 of fiscal year 1970.

```{r divmod}
# Days → fiscal periods
S7::method(chronon_divmod, list(cal_gregorian$day, cal_fiscal$period)) <-
  function(from, to, x) {
    # Scale x to individual days
    scale <- chronon_cardinality(from, cal_gregorian$day(1L))
    days  <- scale * x

    period_size <- chronon_cardinality(to, cal_gregorian$day(1L))  # 28
    list(
      div = days %/% period_size,
      mod = days  %% period_size
    )
  }

# Fiscal periods → days
S7::method(chronon_divmod, list(cal_fiscal$period, cal_gregorian$day)) <-
  function(from, to, x) {
    mult <- chronon_cardinality(from, cal_gregorian$day(1L))  # 28
    list(
      div = x * mult,
      mod = 0L
    )
  }

# Days → fiscal years
S7::method(chronon_divmod, list(cal_gregorian$day, cal_fiscal$year)) <-
  function(from, to, x) {
    scale       <- chronon_cardinality(from, cal_gregorian$day(1L))
    days        <- scale * x
    year_size   <- 13L * 28L  # 364 days per fiscal year
    list(
      div = days %/% year_size,
      mod = days  %% year_size
    )
  }

# Fiscal years → days
S7::method(chronon_divmod, list(cal_fiscal$year, cal_gregorian$day)) <-
  function(from, to, x) {
    mult <- 13L * 28L  # 364
    list(
      div = x * mult,
      mod = 0L
    )
  }

# Quick sanity check - day 28 should be the start of period 2
with(cal_fiscal, chronon_divmod(cal_gregorian$day(1L), period(1L), 28L))
```

## Labels

Three generics control how time units appear as text.

### `time_unit_full()` and `time_unit_abbr()`

These return the singular full name and the short abbreviation for a unit,
respectively. They are used in messages and in tsibble interval displays.

```{r unit-labels}
S7::method(time_unit_full, cal_fiscal$year)   <- function(x) "fiscal year"
S7::method(time_unit_abbr, cal_fiscal$year)   <- function(x) "FY"

S7::method(time_unit_full, cal_fiscal$period) <- function(x) "fiscal period"
S7::method(time_unit_abbr, cal_fiscal$period) <- function(x) "FP"

time_unit_full(cal_fiscal$year(1L))    # "fiscal year"
time_unit_abbr(cal_fiscal$period(1L))  # "FP"
```

### `cyclical_labels()`

`cyclical_labels(granule, cycle, i)` returns human-readable labels for the
0-indexed position `i` of `granule` within `cycle`. This is used when
formatting cyclical time vectors (e.g. "Jan", "Feb", … for months within a
year).

The default method falls back to `paste0(time_unit_abbr(granule), i)`, so you
only need to override it when you want custom labels.

```{r cyclical-labels}
# Fiscal periods within a fiscal year: "FP01", "FP02", …, "FP13"
S7::method(cyclical_labels, list(cal_fiscal$period, cal_fiscal$year)) <-
  function(granule, cycle, i, ...) {
    sprintf("FP%02d", i + 1L)
  }

# Demonstrate the default and the custom override
with(cal_gregorian, cyclical_labels(month(1L),  year(1L),   0:11))
with(cal_isoweek,   cyclical_labels(day(1L),    week(1L),   0:6))
with(cal_fiscal,    cyclical_labels(period(1L), year(1L),   0:12))
```

## Time vector helpers

Once the calendar algebra is in place you can build reusable time-vector
constructors.

### `new_linear_time_fn()`

`new_linear_time_fn(chronon, granules, fallback_calendar)` returns a function
that converts arbitrary input to a linear (non-repeating) time vector. The
`chronon` is the finest resolution stored; `granules` lists coarser units used
for display grouping.

```{r linear-fn}
# Year-period linear time (like yearmonth() but for fiscal periods)
fiscal_period <- new_linear_time_fn(
  chronon           = period(1L),
  granules          = list(year(1L)),
  fallback_calendar = cal_fiscal
)

# Fiscal year only
fiscal_year <- new_linear_time_fn(
  chronon           = year(1L),
  fallback_calendar = cal_fiscal
)

fiscal_period(as.Date("1970-01-01") + 0:3 * 28)
fiscal_year(as.Date("1970-01-01") + 0:2 * 364)
```

### `new_cyclical_time_fn()`

`new_cyclical_time_fn(chronon, cycle, fallback_calendar)` returns a function
that captures the *position within a repeating cycle* - for example the period
number within a fiscal year.

```{r cyclical-fn}
period_of_fiscal_year <- new_cyclical_time_fn(
  chronon           = period(1L),
  cycle             = list(year(1L)),
  fallback_calendar = cal_fiscal
)

# Period number (1–13) for a handful of dates
period_of_fiscal_year(as.Date("1970-01-01") + c(0, 28, 56, 364))
```
